---
title: "HW5"
author: "John Horton"
date: "October 6, 2015"
output: html_document
---

Setup:

```{r}
#Load required packages
require(ggplot2)
require(grid)

#Load diamonds dataframe
data(diamonds)
```

Question 1:

The function below takes a dataframe from the user and returns a new dataframe containing only the numeric columns of the original dataframe. More detail on the function's operation is provided below in the inline comments.

Parameters:
 - dataFrame: the dataframe passed by the user
 
Return:
 - numDataFrame: the dataframe containing only the numeric columns of dataFrame
 
```{r}
Extract_Num <- function(dataFrame){
  #We use the sapply function on dataFrame to isolate only those columns in dataFrame for which 'is.numeric' returns 'TRUE'
  numDataFrame <- dataFrame[sapply(dataFrame,is.numeric)]
  return(numDataFrame)
}

#Tests the function
testNum = Extract_Num(diamonds)
print(diamonds)
print(testNum)
```

Question 2:

The function below takes a dataframe from the user and returns a new dataframe containing the Pearson correlation coefficient for each pair of numeric columns in the original dataframe. More detail on the function's operation is provided below in the inline comments.

Parameters:
 - dataFrame: the dataframe passed by the user
 
 Return:
  - corrFrame: the dataframe containing the Pearson correlation coefficient for each pair of numeric columns in dataFrame
  
```{r}
Column_Corr <- function(dataFrame){
  #We first call Extract_Num to isolate the numeric columns of dataFrame
  dataFrame <- Extract_Num(dataFrame)
  #The vector corrTitle, which will hold the titles of each pair of numeric columns, is initialized as an empty vector
  #Each value in this vector will be of the form 'col1-col2', where col1 and col2 are the titles of column 1 and column 2
  corrTitle <- c()
  #The vector corrValue, which will hold the Pearson correlation coefficient for each pair of numeric columns, is initialized as an empty vector
  corrValue <- c()
  #The index variable k is initialized to 1; k corresponds to the total number of non-repeating pairs of numeric columns in dataFrame
  k = 1
  #We loop through each column in dataFrame so that we may take the correlation between that column and the next
  for(i in 1:(ncol(dataFrame)-1)){
    #We start our next loop at i+1, so that we do not repeat pairs of columns
    #Starting our loop at i+1 also avoids taking unneccessary correlations between a column and itself, i.e. we do not take the correlation between carat and carat, as that is obviously 1
    for(j in (i+1):(ncol(dataFrame))){
      #We take the title of the current pair of columns
      corrTitle[k] <- paste(colnames(dataFrame)[i],"-",colnames(dataFrame)[j],sep="")
      #We take the correlation of the current pair of columns
      corrValue[k] <- cor(dataFrame[,i],dataFrame[,j],method="pearson")
      #We increase k by 1, as we now consider the next pair
      k <- k+1
    }
  }
  #We initialize the return variable to an empty dataframe with one row for each pair of columns taken above
  corrFrame <- data.frame(matrix(ncol=0,nrow=k-1))
  #We insert two new columns to our return dataframe, one for the title of each pair of columns, another for the correlation of each pair of columns
  corrFrame["Columns"] <- corrTitle
  corrFrame["Correlation"] <- corrValue
  return(corrFrame)
}

#Tests the function
testCorr = Column_Corr(diamonds)
#print(diamonds)
print(testCorr)

```

Question 3:

This function takes a dataframe from the user and returns scatter plots for each pair of numeric columns in that dataframe. The title of each scatter plot will contain the variables contained in that plot and the Pearson correlation coefficient for that pair of variables. Plots are returned in a list.

Parameters:
 - dataFrame: the dataframe passed by the user
 
Return:
 - corrPlots: the list containing a plot for each pair of numeric columns in dataFrame
 
```{r}
Corr_Plot <- function(dataFrame){
  #We first call Extract_Num to isolate the numeric columns of dataFrame
  dataFrame <- Extract_Num(dataFrame)
  #The list corrPlots, which will hold the scatter plot for each pair of numeric columns, is initialized as an empty list
  corrPlots <- list()
  #The index variable k is initialized to 1; k corresponds to the total number of non-repeating pairs of numeric columns in dataFrame
  k <- 1
    #We loop through each column in dataFrame so that we may plot that column and the next, and take their correlation
  for(i in 1:(ncol(dataFrame)-1)){
    #We start our next loop at i+1, so that we do not repeat pairs of columns
    #Starting our loop at i+1 also avoids taking unneccessary correlations between a column and itself, i.e. we do not take the correlation between carat and carat, as that is obviously 1
    #We save even more time from this change here than in Column_Corr as we avoid producing unneccessary plots, important here because producing each plot takes a noticeable amount of computing time 
    for(j in (i+1):(ncol(dataFrame))){
      #The dataframe plotFrame is created to hold just the data that we wish to plot in this iteration of the loop
      plotFrame <- data.frame(dataFrame[,i],dataFrame[,j])
      #xLabel stores the name of the column to be plotted on the x axis, so we may avoid calling a function inside our aes call below
      xLabel <- colnames(dataFrame)[i]
      #yLabel stores the name of the column to be plotted on the y axis, so we may avoid calling a function inside our aes call below
      yLabel <- colnames(dataFrame)[j]
      #We set the column names in plotFrame to xLabel and yLabel, so that ggplot will recognize them
      colnames(plotFrame) <- c(xLabel,yLabel)
      #xyCorr stores the correlation between the two columns contained in this plot, which will be added to the title of this plot as per the question prompt
      xyCorr <- cor(dataFrame[,i],dataFrame[,j],method="pearson")
      #Here we make the appropriate ggplot calls to produce our scatter plot
      #We feed ggplot the dataframe plotFrame, call aes_string to set the axes, call geom_point to produce a scatter plot rather than another type of graph, and call ggtitle to give the plot the desired title
      #Each plot is stored in our list of plots, which is returned once we have gone through both loops
      corrPlots[[k]] <- (ggplot(plotFrame,aes_string(x=xLabel,y=yLabel))+geom_point()+ggtitle(paste("Correlation between",xLabel,"and",yLabel,":",xyCorr)))
      #We increase k by 1, as we now consider the next pair
      k <- k+1
    }
  }
  return(corrPlots)
}

#Tests the function
testPlots = Corr_Plot(diamonds)
print(testPlots)
```
